---
title: "broad-sw-new"
format: html
---

Does elephant seal movement change in the presence of sperm whale clicks?

```{r}
#| label: setup
#| message: false

library(aniMotum)
library(tidyverse)
library(terra)
library(tidyterra)
library(argosfilter)

theme_set(theme_bw())
```

Formatting location data (a mix of GPS and Argos). Based on Storrie et al., 2025 and the Wildlife Computers documentation.

```{r}
track <- read_csv(here::here("data/255226-gps-1/H391-255226-1-Locations.csv"),
                  show_col_types = FALSE,
                  locale = locale(tz = "UTC")) %>% 
  #remove Argos locations with no associate ellipse error values and GPS locations with fewer than 4 sattelites. 
  filter(!(Type == "Argos" & is.na(`Error Ellipse Orientation`)), 
         !(Type == "FastGPS" & suppressWarnings(as.numeric(Quality) < 4))) %>% 
#Warning because as.numeric() converts the Argos quality estimates to NAs
  mutate(Date = parse_date_time(Date, 
                                orders = c("H:M:S.OS d-b-Y", 
                                           "H:M:S d-b-Y")))

#Remove locations that would require unrealistic swimming speeds (>3m/s)
track_filtered <- track %>% 
  arrange(Date) %>% 
  mutate(filter_flag = sdafilter(
    lat = Latitude, 
    lon = Longitude, 
    dtime = Date, 
    lc = Quality,
    vmax = 3
  )) %>% 
  filter(filter_flag == "not") %>% #keep only realistic locations 
  ungroup()
```

Read in recording details and clip track to recording time.

```{r}
#| label: acou-data

#read in the recording details
#wav_timestamps.csv is the output of get_wav_timestamps.R. 
recording <- read.csv(here::here("data/wav_timestamps.csv")) %>% 
  mutate(start = ymd_hms(start_time, tz = "UTC"),
         end = ymd_hms(end_time, tz = "UTC")) %>% 
  drop_na(end)

#filter the track to only when the acoustic tag was recording
track_overlap <- filter(track_filtered, between(Date, 
                                       first(recording$start),
                                       last(recording$end)))

#on average, during the recording period, how often did we get a filtered location?
avg_loc_timing <- track_overlap %>% 
  mutate(interval = as.numeric(difftime(Date, 
                                        lag(Date), 
                                        units = "hours"))) %>% 
  summarize(mean_interval_hrs = mean(interval, na.rm = TRUE)) %>% 
  round(1)
```

On average, we got a filtered location every `r avg_loc_timing$mean_interval_hrs` hours.

```{r}
#| label: plot-track-initial

track_overlap %>%
  ggplot(aes(Longitude, Latitude)) +
  geom_path() 

```

Format track data for `aniMotum`.

```{r}
#| label: format-track-animotum

track_formatted <- track_overlap %>% 
  mutate(id = c("H391"), 
         `Error Radius` = if_else(Type == "FastGPS", 50, `Error Radius`), 
         Quality = if_else(Type == "FastGPS", "G", Quality)) %>% 
  format_data(date = "Date", 
              coord = c("Longitude", "Latitude"), 
              lc = "Quality") %>% 
  as.data.frame()
```

```{r}
#| label: fit-initial-ssm

fit <- fit_ssm(track_formatted, 
               vmax = 3, #max travel rate 
               model = "mp", #fits correlated random walk
               time.step = 6,
               control = ssm_control(verbose = 0))

summary(fit)
p1 <- plot(fit, type = 3)$H391
p2 <- ggplot(grab(fit, as_sf = TRUE)) + 
  geom_sf(aes(color = g)) +
  scale_color_gradient2(midpoint = 0.5, limits = c(0, 1))

p1
```

Now read in the manual detections of sperm whales.

```{r}
source(here::here("R/broad_funcs.R"))

sw_valid <- read_csv(here::here("data/sw/SWCombBouts_Validated.csv"), 
                     show_col_types = FALSE) %>% 
  mutate(start= as.POSIXct(start, tz = "UTC"), 
         end = as.POSIXct(end, tz = "UTC"), 
         duration = end - start) %>% 
  select(-notes)

window_hr <- 6
acou_agg_valid <- tibble(date = seq(first(recording$start), 
                                    last(recording$end), 
                                    by = window_hr * 3600)) %>% 
  mutate(date2 = date - window_hr * 3600,
         SW = map_dbl(date, \(d) sum_odon_exposure(data = sw_valid, 
                                                   win_start = d, 
                                                   win_hr = window_hr)),
         SW_lag = lag(SW), 
         SW_lead = lead(SW), 
         SW_std = (SW - mean(SW)) / sd(SW),
         id = "H391")

total_sw_hrs <- round(sum(as.numeric(sw_valid$duration)), 1)
```

The seal recorded `r total_sw_hrs` of sperm whales (out of 740 total hours of recording - about `r round(total_sw_hrs/740 * 100, 1)`% of the time).

Plot the track with SW bouts.

```{r}
#| label: track_bouts

track_overlap_sw <- track_overlap %>% 
  left_join(
    acou_agg_valid %>% select(date2, SW),
    join_by(closest(Date >= date2))
  ) %>% 
  rename(sw = SW) %>%
  ggplot(aes(Longitude, Latitude)) +
  geom_path() +
  geom_point(aes(color = sw)) +
  scale_color_viridis_c(option = "inferno")

```

Plot movement persistence with SW presence (just for visualization purposes).

```{r}
p_sw <- plot(fit, type = 3)$H391 +
  geom_point(aes(date, SW / max(SW)), acou_agg_valid) +
  scale_y_continuous(sec.axis = sec_axis(~ . * max(acou_agg_valid$SW), name = "SW (hours)"))

cowplot::plot_grid(track_overlap_sw, p_sw, ncol = 2, rel_widths = c(2, 5))
```

Look at the change in the mean of SW detections over time using a breakpoint model.

```{r}
#| label: sw-change

library(changepoint)
sw_cpt <- cpt.mean(acou_agg_valid$SW, method = "AMOC")
breakpoint_date <- acou_agg_valid$date[sw_cpt@cpts[1]]
sw_segs <- tibble(
  seg_start = acou_agg_valid$date[c(1, sw_cpt@cpts[1])],
  seg_end = acou_agg_valid$date[c(sw_cpt@cpts[1], nrow(acou_agg_valid))],
  seg_mean = sw_cpt@param.est$mean
)

ggplot(acou_agg_valid, aes(x = date, y = SW)) +
  geom_point(alpha = 0.4) +
  geom_segment(aes(x = seg_start, xend = seg_end, y = seg_mean, yend = seg_mean),
               sw_segs,
               linetype = "dashed",
               color = "hotpink",
               linewidth = 1.2) +
  geom_vline(xintercept = breakpoint_date, 
             linetype = "dotted", 
             color = "hotpink", 
             linewidth = 1.2) +
  labs(x = NULL, y = "SW detections") +
  theme_classic()
```

Compare to a breakpoint model for the mixed layer temperature.

```{r}
#| label: mlt-mod

#mixed layer temperature - using average MLT and minimum temperature from this file
mlt <- read_csv(here::here("data/23A1302/out-MixLayer.csv"))
mlt <- mlt %>% 
  mutate(date = as.POSIXct(Date, 
                        format = "%H:%M:%S %d-%b-%Y", 
                        tz = "UTC"),
         date_num = as.numeric(date - date[1], unit = "days"),
         id = "H391") %>% 
   filter(between(date, 
                  first(acou_agg_valid$date2), 
                  last(acou_agg_valid$date)))

# There's one clear outlier in MLT temperature, remove it and interpolate
outlier_idx <- which(mlt$MLTave > 14)
mlt$MLTave[outlier_idx] <- mean(mlt$MLTave[outlier_idx + c(-1, 1)])
mlt$MLTave_std <- (mlt$MLTave - mean(mlt$MLTave)) / sd(mlt$MLTave)

mlt_cpt <- cpt.reg(cbind(mlt$MLTave, 1, mlt$date_num), method = "AMOC")
mlt_cpt_date <- mlt$date[mlt_cpt@cpts[1]]
b <- param.est(mlt_cpt)$beta
mlt_segs <- tibble(
  date = mlt$date[c(1, mlt_cpt@cpts[1], mlt_cpt@cpts[1], nrow(mlt))],
  date_num = as.numeric(date - mlt$date[1], unit = "days"),
  seg_num = c(1, 1, 2, 2),
  MLTave = c(b[1, 1] + b[1, 2] * date_num[1],
             b[1, 1] + b[1, 2] * date_num[2],
             b[2, 1] + b[2, 2] * date_num[3],
             b[2, 1] + b[2, 2] * date_num[4])
)

ggplot(mlt, aes(x = date, y = MLTave)) +
  geom_point(alpha = 0.4) +
  geom_line(aes(group = seg_num), 
            mlt_segs, 
            linetype = "dashed",
            color = "firebrick",
            linewidth = 1.2) +
  geom_vline(xintercept = mlt_cpt_date, 
             linetype = "dotted", 
             color = "firebrick", 
             linewidth = 1.2) +
  labs(x = NULL, y = "Average Mixed layer temp") +
  theme_classic()
```

```{r}
p1 +
  geom_vline(xintercept = breakpoint_date, 
             linetype = "dashed", 
             color = "hotpink",
             linewidth = 1.2) +
  geom_vline(xintercept = mlt_cpt_date, 
             linetype = "dashed", 
             color = "firebrick",
             linewidth = 1.2)

```
