# Broad scale

Using the `aniMotum` package as a way to fit a continuous HMM to the seal GPS data.

```{r}
#| label: setup
#| message: false

library(aniMotum)
library(tidyverse)
library(terra)
library(tidyterra)

theme_set(theme_bw())
```

Loading in our GPS data and formatting it to work with `aniMotum`:

```{r}
#| label: track-data

track <- read_csv(here::here("data/seal_locations.csv"),
                  show_col_types = FALSE,
                  locale = locale(tz = "UTC")) %>% 
  mutate(lc = c("G"), 
         id = c("H391")) %>% 
  format_data(date = "Date_es", coord = c("Longitude", "Latitude")) %>% 
  as.data.frame()

```

Load in acoustic data and aggregate it at 6-hour scale

```{r}
#| label: acou-data

source(here::here("R/broad_funcs.R"))

#read in the recording details
#wav_timestamps.csv is the output of the function get_wav_timestamps.R. 
recording <- read.csv(here::here("data/wav_timestamps.csv")) %>% 
                        mutate(start = ymd_hms(start_time, tz = "UTC"),
                               end = ymd_hms(end_time, tz = "UTC")) %>% 
                        drop_na(end)

#read in sperm whale data 
sw_valid <- read_csv(here::here("data/sw/SWCombBouts_Validated.csv"), 
                     show_col_types = FALSE) %>% 
  mutate(start= as.POSIXct(start, tz = "UTC"), 
         end = as.POSIXct(end, tz = "UTC"), 
         duration = end - start) %>% 
  select(-notes)

window_hr <- 6

acou_agg_valid <- tibble(date = seq(first(recording$start), 
                                    last(recording$end), 
                                    by = window_hr * 3600)) %>% 
  mutate(date2 = date - window_hr * 3600,
         SW = map_dbl(date, \(d) sum_odon_exposure(data = sw_valid, 
                                                   win_start = d, 
                                                   win_hr = window_hr)),
         SW_lag = lag(SW), 
         SW_lead = lead(SW), 
         SW_std = (SW - mean(SW)) / sd(SW),
         id = "H391")

# acou_agg_valid <- tibble(date = seq(first(track$date), 
#                                     last(track$date), 
#                                     by = window_hr * 3600)) %>% 
#   mutate(date2 = date - window_hr * 3600,
#          SW = map_dbl(date, \(d) sum_odon_exposure(data = sw_valid, d)),
#          SW_lag = lag(SW), 
#          SW_lead = lead(SW), 
#          SW_std = (SW - mean(SW)) / sd(SW),
#          id = "H391")

```

The seal recorded `sum(sw_valid$duration)` of sperm whales (out of 740 total hours of recording - about `sum(sw_valid$duration)/740`%.

Trim spatial, acoustic data to audio coverage

```{r}
#| label: trim-data

track_overlap <- filter(track, between(date, 
                                       first(acou_agg_valid$date2), 
                                       last(acou_agg_valid$date)))

```

Sanity check: plot track and sperm whale bouts.

```{r}
#| label: track_bouts

track_overlap %>% 
  mutate(sw = map_dbl(date, \(d) acou_agg_valid$SW[d >= acou_agg_valid$date2 & d < acou_agg_valid$date])) %>%
  ggplot(aes(lon, lat)) +
  geom_path() +
  geom_point(aes(color = sw)) +
  scale_color_viridis_c(option = "inferno")
  

```

Fit model

```{r}
#| label: fit-mod

fit <- fit_ssm(track_overlap, 
               vmax = 3, #max travel rate 
               model = "mp", #fits correlated random walk
               time.step = select(acou_agg_valid, id, date),
               control = ssm_control(verbose = 0))

#this isn't working because we changed back to standard animotum 
# fit_sw <- fit_ssm(track_overlap,
#                vmax = 3, #max travel rate
#                model = "mp", #fits correlated random walk
#                time.step = select(acou_agg_valid, id, date), #6 hour steps
#                control = ssm_control(verbose = 0),
#                env_formula = ~ SW,
#                env_data = select(acou_agg_valid, id, date, SW))

```

```{r}
#| label: summ-mod

fit$ssm$H391$par
# est <- fit$ssm$H391$par["beta", 1]
# se <- fit$ssm$H391$par["beta", 2]
# pnorm(est, mean = 0, sd = se, lower.tail = TRUE)

```

Instead of jointly estimating the random walk of g in the model, estimate it after fitting g.

```{r}
#| label: fit-g

logit <- function(x) log(x / (1 - x))
inv_logit <- function(x) exp(x) / (1 + exp(x))

g_df <- fit$ssm$H391$predicted %>% 
  mutate(SW = acou_agg_valid$SW,
         logit_g = logit(g),
         logit_g_lag = lag(logit_g)) %>% 
  slice(-1)
g_rw <- lm(logit_g ~ 0 + offset(logit_g_lag) + SW, g_df)
summary(g_rw)


```

```{r}
#| label: plot-mod

p <- plot(fit, type = 3)$H391
p +
  geom_point(aes(date, SW / max(SW)), acou_agg_valid) +
  scale_y_continuous(sec.axis = sec_axis(~ . * max(acou_agg_valid$SW), name = "SW (hours)"))

```

```{r}
#| label: plot-map

ggplot(grab(fit, as_sf = TRUE)) + 
  geom_sf(aes(color = g)) +
  scale_color_gradient2(midpoint = 0.5, limits = c(0, 1))
```

## Alternative hypotheses:

### 1. Luck

Maybe sperm whales and elephant seals both just happened to converge on the patch. If that's true, then adding a 6 hour *lag time* or *lead time* to the model should have similar results.

```{r}
#| label: lead-mod

# No need to re-fit the model anymore
# fit_lead <- fit_ssm(track, 
#                vmax = 3, #max travel rate 
#                model = "mp", #fits correlated random walk
#                time.step = select(acou_agg_valid, id, date), #12 hour steps
#                control = ssm_control(verbose = 0), 
#                env_formula = ~ SW_lead,
#                env_data = acou_agg_valid)

#get lead results

# fit_lead$ssm$H391$par
# est_lead <- fit_lead$ssm$H391$par["beta", 1]
# se_lead <- fit_lead$ssm$H391$par["beta", 2]
# pnorm(est_lead, mean = 0, sd = se_lead, lower.tail = TRUE)

g_df_lead <- fit$ssm$H391$predicted %>% 
  mutate(SW_lead = acou_agg_valid$SW_lead,
         logit_g = logit(g),
         logit_g_lag = lag(logit_g)) %>% 
  slice(-1)
g_rw_lead <- lm(logit_g ~ 0 + offset(logit_g_lag) + SW_lead, 
                g_df_lead)
summary(g_rw_lead)
```

```{r}
#| label: lag-mod

fit_lag <- fit_ssm(track, 
               vmax = 3, #max travel rate 
               model = "mp", #fits correlated random walk
               time.step = select(acou_agg_valid, id, date), #12 hour steps
               control = ssm_control(verbose = 0), 
               env_formula = ~ SW_lag,
               env_data = acou_agg_valid)

#get lag results
fit_lag$ssm$H391$par
est_lag <- fit_lag$ssm$H391$par["beta", 1]
se_lag <- fit_lag$ssm$H391$par["beta", 2]
pnorm(est_lag, mean = 0, sd = se_lag, lower.tail = TRUE)

g_df_lag <- fit$ssm$H391$predicted %>% 
  mutate(SW_lag = acou_agg_valid$SW_lag,
         logit_g = logit(g),
         logit_g_lag = lag(logit_g)) %>% 
  slice(-1)
g_rw_lag <- lm(logit_g ~ 0 + offset(logit_g_lag) + SW_lag, g_df_lag)
summary(g_rw_lag)
```

A model fitted to the lag/lead values are not at *all* significant (.5-.6 one-tailed p-test), while the model fitted to the actual values is (0.03, 1 tailed p test).

### 2. Memory

Maybe H391 has been to that patch before and knew that there was good food there, and she didn't actually use the sperm whale cue at all.

We don't know what she does every year, but we do know that for her two other instrumented trips, she traveled offshore - so this isn't just "her spot."

```{r}
#| label: memory-plots

#plotting her previous trips
library(ggOceanMaps)
library(terra)
library(tidyterra)
theme_set(theme_bw())

pb25 <- read_csv(here::here("data/seal_locations.csv"),
                          show_col_types = FALSE) %>% 
  rename(Date = Date_es) %>% 
  mutate(trip = "PB") %>% 
  vect(geom = c("Longitude", "Latitude"), 
       crs = "epsg:4326")

pb_track <- as.lines(pb25)

pm25 <- read_csv(here::here("data/MoreDeployments/H391-PM2025/265936-Locations.csv"), 
                 show_col_types = FALSE) %>% 
  mutate(trip = "PM") %>% 
  select(Date, Latitude, Longitude, trip) %>% 
  vect(geom = c("Longitude", "Latitude"), 
       crs = "epsg:4326")

pm_track <- as.lines(pm25)

rod <- read_csv(here::here("data/MoreDeployments/H391-RoD/2023228_240257_aniMotum_crw_RSB.csv"), 
                show_col_types = FALSE) %>% 
  rename(Date = date, 
         Longitude = lon, 
         Latitude = lat) %>% 
  mutate(trip = "ROD") %>% 
  select(Date, Latitude, Longitude, trip) %>% 
  vect(geom = c("Longitude", "Latitude"), 
       crs = "epsg:4326")

rod_track <- as.lines(rod)

basemap(limits = c(-163, -121, 33, 58),
        bathymetry = TRUE,
        bathy.style = "rcb") +
  geom_spatvector(data = pm_track, color = "darkblue", linewidth = 1) +
  geom_spatvector(data = rod_track, color = "lightskyblue", linewidth = 1) +
  geom_spatvector(data = pb_track, color = "gold", linewidth = 1) +
  xlab("Longitude") + 
  ylab("Latitude") +
  theme(legend.position = "none")

```

Both of the other deployments were longer trips, but many seals do go offshore on the shorter trips as well. The trip with acoustic recordings is plotted in gold.

### 3. Other environmental cues

Maybe the seal wasn't using sperm whales at all, but their calls happened to coincide with another environmental cue, like SST, temp at depth, or FTLE. We could use the tag's temperature sensor and FTLE measurements to fit our same SSM with oceanographic covariates instead of sperm whale presence and see if the fit is better or worse than with SW.

Let's try it with average mixed layer temperature, as measured by the tag.

```{r}
#mixed layer temperature - using average MLT and minimum temperature from this file
mlt <- read_csv(here::here("data/23A1302/out-MixLayer.csv"))
mlt <- mlt %>% 
  mutate(date = as.POSIXct(Date, 
                        format = "%H:%M:%S %d-%b-%Y", 
                        tz = "UTC"), 
         id = "H391") %>% 
   filter(between(date, 
                  first(acou_agg_valid$date2), 
                  last(acou_agg_valid$date)))

# There's one clear outlier in MLT temperature, remove it and interpolate
outlier_idx <- which(mlt$MLTave > 14)
mlt$MLTave[outlier_idx] <- mean(mlt$MLTave[outlier_idx + c(-1, 1)])
mlt$MLTave_std <- (mlt$MLTave - mean(mlt$MLTave)) / sd(mlt$MLTave)
```

```{r}
#| label: mlt-mod

fit_sst <- fit_ssm(track, 
               vmax = 3, #max travel rate 
               model = "mp", #fits correlated random walk
               time.step = select(mlt, id, date), #6 hour steps
               env_formula = ~ MLTave_std,
               env_data = select(mlt, id, date, MLTave_std), 
               control = ssm_control(verbose = 0), 
)

fit_sst$ssm$H391$par
est <- fit_sst$ssm$H391$par["beta", 1]
se <- fit_sst$ssm$H391$par["beta", 2]
pnorm(est, mean = 0, sd = se, lower.tail = TRUE)

g_df_mlt <- fit$ssm$H391$predicted %>% 
  mutate(mlt = mlt$MLTave,
         logit_g = logit(g),
         logit_g_lag = lag(logit_g)) %>% 
  slice(-1)
g_rw_mlt <- lm(logit_g ~ 0 + offset(logit_g_lag) + mlt, g_df_mlt)
summary(g_rw_mlt)
```

```{r}
p <- plot(fit_sst, type = 3)$H391
p +
  geom_point(aes(date, MLTave / max(MLTave)), mlt) +
  scale_y_continuous(sec.axis = sec_axis(~ . * max(mlt$MLTave), name = "MLT (degC)"))
```

The mixed layer average temperature model isn't significant. Check the fit of this model compared to the SW model:

```{r}
fit$ssm$H391$AICc
fit_sst$ssm$H391$AICc
```

SW model is a better fit.

#### Next step: the fine scale.

We will fit a hidden markov model to the seal's diving behavior and allow the properties of the sperm whale clicks (received level and peak frequency, both associated with how far away the whales are) to influence her transition probabilities.
