---
title: "broad-sw-new"
format: html
---

Does elephant seal movement change in the presence of sperm whale clicks?

```{r}
#| label: setup
#| message: false

library(aniMotum)
library(tidyverse)
library(terra)
library(tidyterra)
library(argosfilter)
library(marmap)      # ETOPO1 bathymetry
library(metR)        # geom_contour_fill 
library(rnaturalearth)
library(rnaturalearthdata)

theme_set(theme_bw())
```

Formatting location data (a mix of GPS and Argos). Based on Storrie et al., 2025 and the Wildlife Computers documentation.

```{r}
track <- read_csv(here::here("data/255226-gps-1/H391-255226-1-Locations.csv"),
                  show_col_types = FALSE,
                  locale = locale(tz = "UTC")) %>% 
  #remove Argos locations with no associate ellipse error values and GPS locations with fewer than 4 sattelites. 
  filter(!(Type == "Argos" & is.na(`Error Ellipse Orientation`)), 
         !(Type == "FastGPS" & suppressWarnings(as.numeric(Quality) < 4))) %>% 
#Warning because as.numeric() converts the Argos quality estimates to NAs
  mutate(Date = parse_date_time(Date, 
                                orders = c("H:M:S.OS d-b-Y", 
                                           "H:M:S d-b-Y")))

#Remove locations that would require unrealistic swimming speeds (>3m/s)
track_filtered <- track %>% 
  arrange(Date) %>% 
  mutate(filter_flag = sdafilter(
    lat = Latitude, 
    lon = Longitude, 
    dtime = Date, 
    lc = Quality,
    vmax = 3
  )) %>% 
  filter(filter_flag == "not") %>% #keep only realistic locations 
  ungroup()
```

Read in recording details and clip track to recording time.

```{r}
#| label: acou-data

#read in the recording details
#wav_timestamps.csv is the output of get_wav_timestamps.R. 
recording <- read.csv(here::here("data/wav_timestamps.csv")) %>% 
  mutate(start = ymd_hms(start_time, tz = "UTC"),
         end = ymd_hms(end_time, tz = "UTC")) %>% 
  drop_na(end)

#filter the track to only when the acoustic tag was recording
track_overlap <- filter(track_filtered, between(Date, 
                                       first(recording$start),
                                       last(recording$end)))

#on average, during the recording period, how often did we get a filtered location?
avg_loc_timing <- track_overlap %>% 
  mutate(interval = as.numeric(difftime(Date, 
                                        lag(Date), 
                                        units = "hours"))) %>% 
  summarize(mean_interval_hrs = mean(interval, na.rm = TRUE)) %>% 
  round(1)
```

On average, we got a filtered location every `r avg_loc_timing$mean_interval_hrs` hours.

```{r}
#| label: plot-track-initial

bathy <- getNOAA.bathy(
  lon1 = -126, lon2 = -122, 
  lat1 = 35, lat2 = 45, 
  resolution = 2, 
  keep = TRUE
)
bathy_df <- fortify(bathy)
isobaths <- c(-200, -500, -1000, -2000, -4000) 

land <- ne_countries(scale = "medium", returnclass = "sf")
states <- ne_states(country = "united states of america", returnclass = "sf") 

track_map <- ggplot() + 
  #bathymetry
  geom_raster(data = filter(bathy_df, z <= 0), 
              aes(x, y, fill = z), 
              interpolate = TRUE) + 
  scale_fill_gradientn(
    colours  = c("#08306b", "#2171b5", "#6baed6", "#c6dbef"),
    values   = scales::rescale(c(-6000, -2000, -200, 0)),
    name     = "Depth (m)",
    na.value = "grey80"
  ) +
  # Land + state borders
  geom_sf(data = land,
          fill = "grey75",
          colour = NA) +
  geom_sf(data = states, 
          fill = NA, 
          colour = "grey60", 
          linewidth = 0.2) +  
  coord_sf(
    xlim   = c(-126, -122),
    ylim   = c(37, 44),
    expand = FALSE
  ) + 
  scale_x_continuous(breaks = seq(-126, -122, by = 2)) +
  # Isobath contours
  geom_contour(
    data    = bathy_df,
    aes(x, y, z = z),
    breaks  = isobaths,
    colour  = "white", linewidth = 0.25, alpha = 0.6
  ) + 
  #Entire track
  geom_path(data = track_filtered, 
            aes(Longitude, Latitude), 
            color = "goldenrod", 
            alpha = .5) + 
  labs(x = "Longitude", 
       y = "Latitude") +
  # Track with recording
  geom_path(data = track_overlap, 
            aes(Longitude, Latitude), 
            color = "gold") + 
  labs(x = "Longitude", 
       y = "Latitude") + 
  theme_minimal()
track_map
```

Format track data for `aniMotum`.

```{r}
#| label: format-track-animotum

track_formatted <- track_overlap %>% 
  mutate(id = c("H391"), 
         `Error Radius` = if_else(Type == "FastGPS", 50, `Error Radius`), 
         Quality = if_else(Type == "FastGPS", "G", Quality)) %>% 
  format_data(date = "Date", 
              coord = c("Longitude", "Latitude"), 
              lc = "Quality") %>% 
  as.data.frame()
```

```{r}
#| label: fit-initial-ssm

fit <- fit_ssm(track_formatted, 
               vmax = 3, #max travel rate 
               model = "mp", #fits correlated random walk
               time.step = 3,
               control = ssm_control(verbose = 0))

# summary(fit)
p1 <- plot(fit, type = 3)$H391
p2 <- ggplot(grab(fit, as_sf = TRUE)) + 
  geom_sf(aes(color = g)) +
  scale_color_gradient2(midpoint = 0.5, limits = c(0, 1))

p1
```

Now read in the manual detections of sperm whales.

```{r}
source(here::here("R/broad_funcs.R"))

sw_valid <- read_csv(here::here("data/sw/SWCombBouts_Validated.csv"), 
                     show_col_types = FALSE) %>% 
  mutate(start= as.POSIXct(start, tz = "UTC"), 
         end = as.POSIXct(end, tz = "UTC"), 
         duration = end - start) %>% 
  select(-notes)

window_hr <- 3
acou_agg_valid <- tibble(date = seq(first(recording$start), 
                                    last(recording$end), 
                                    by = window_hr * 3600)) %>% 
  mutate(date2 = date - window_hr * 3600,
         SW = map_dbl(date, \(d) sum_odon_exposure(data = sw_valid, 
                                                   win_start = d, 
                                                   win_hr = window_hr)),
         SW_lag = lag(SW), 
         SW_lead = lead(SW), 
         SW_std = (SW - mean(SW)) / sd(SW),
         id = "H391")

total_sw_hrs <- round(sum(as.numeric(sw_valid$duration)), 1)
```

The seal recorded `r total_sw_hrs` of sperm whales (out of 740 total hours of recording - about `r round(total_sw_hrs/740 * 100, 1)`% of the time).

Plot the track with SW bouts.

```{r}
#| label: track_bouts

track_overlap_sw <- track_overlap %>% 
  left_join(
    acou_agg_valid %>% select(date2, SW),
    join_by(closest(Date >= date2))
  ) %>% 
  rename(sw = SW) 

sw_track_map <- track_map + 
  geom_point(data = track_overlap_sw, 
             aes(Longitude, Latitude, size = sw), 
             color = "hotpink", 
             alpha = .3) + 
  scale_size_area(name = "Hours of SW", 
                  max_size = 4, 
                  breaks = c(0, 2, 4, 6))

sw_track_map
```

Plot movement persistence with SW presence (just for visualization purposes).

```{r}
p_sw <- plot(fit, type = 3)$H391 +
  geom_point(aes(date, SW / max(SW)), acou_agg_valid) +
  scale_y_continuous(sec.axis = sec_axis(~ . * max(acou_agg_valid$SW), name = "SW (hours)"))

cowplot::plot_grid(track_overlap_sw, p_sw, ncol = 2, rel_widths = c(2, 5))
```

Look at the change in the mean of SW detections over time using a breakpoint model.

```{r}
#| label: sw-change

library(changepoint)
sw_cpt <- cpt.mean(acou_agg_valid$SW, method = "AMOC")
breakpoint_date <- acou_agg_valid$date[sw_cpt@cpts[1]]
sw_segs <- tibble(
  seg_start = acou_agg_valid$date[c(1, sw_cpt@cpts[1])],
  seg_end = acou_agg_valid$date[c(sw_cpt@cpts[1], nrow(acou_agg_valid))],
  seg_mean = sw_cpt@param.est$mean
)

sw_segs$seg_end - sw_segs$seg_start
sum(acou_agg_valid$date < breakpoint_date)
sum(acou_agg_valid$date > breakpoint_date)

sw_cpt_plot <- ggplot(acou_agg_valid, aes(x = date, y = SW)) +
  geom_point(alpha = 0.4) +
  geom_segment(aes(x = seg_start, xend = seg_end, y = seg_mean, yend = seg_mean),
               sw_segs,
               linetype = "dashed",
               color = "hotpink",
               linewidth = 1.2) +
  geom_vline(xintercept = breakpoint_date, 
             linetype = "dotted", 
             color = "hotpink", 
             linewidth = 1.2) +
  labs(x = NULL, y = "SW detections") +
  theme_classic()

sw_cpt_plot
```

Compare to a breakpoint model for the mixed layer temperature.

```{r}
#| label: mlt-mod

#mixed layer temperature - using average MLT and minimum temperature from this file
mlt <- read_csv(here::here("data/23A1302/out-MixLayer.csv"))
mlt <- mlt %>% 
  mutate(date = as.POSIXct(Date, 
                        format = "%H:%M:%S %d-%b-%Y", 
                        tz = "UTC"),
         date_num = as.numeric(date - date[1], unit = "days"),
         id = "H391") %>% 
   filter(between(date, 
                  first(acou_agg_valid$date2), 
                  last(acou_agg_valid$date)))

# There's one clear outlier in MLT temperature, remove it and interpolate
outlier_idx <- which(mlt$MLTave > 14)
mlt$MLTave[outlier_idx] <- mean(mlt$MLTave[outlier_idx + c(-1, 1)])
mlt$MLTave_std <- (mlt$MLTave - mean(mlt$MLTave)) / sd(mlt$MLTave)

mlt_cpt <- cpt.reg(cbind(mlt$MLTave, 1, mlt$date_num), method = "AMOC")
mlt_cpt_date <- mlt$date[mlt_cpt@cpts[1]]
b <- param.est(mlt_cpt)$beta
mlt_segs <- tibble(
  date = mlt$date[c(1, mlt_cpt@cpts[1], mlt_cpt@cpts[1], nrow(mlt))],
  date_num = as.numeric(date - mlt$date[1], unit = "days"),
  seg_num = c(1, 1, 2, 2),
  MLTave = c(b[1, 1] + b[1, 2] * date_num[1],
             b[1, 1] + b[1, 2] * date_num[2],
             b[2, 1] + b[2, 2] * date_num[3],
             b[2, 1] + b[2, 2] * date_num[4])
)

mlt_cpt_date - mlt_segs$date[1]
mlt_cpt_date - mlt_segs$date[4]
sum(mlt$date < mlt_cpt_date)
sum(mlt$date > mlt_cpt_date)

mlt_cpt_plot <- ggplot(mlt, aes(x = date, y = MLTave)) +
  geom_point(alpha = 0.4) +
  geom_line(aes(group = seg_num), 
            mlt_segs, 
            linetype = "dashed",
            color = "firebrick",
            linewidth = 1.2) +
  geom_vline(xintercept = mlt_cpt_date, 
             linetype = "dotted", 
             color = "firebrick", 
             linewidth = 1.2) +
  labs(x = NULL, y = "Average Mixed layer temp") +
  theme_classic()

cowplot::plot_grid(sw_cpt_plot, mlt_cpt_plot, nrow = 2)
```

```{r}
p_mp_cpt <- p1 +
  geom_vline(xintercept = breakpoint_date, 
             linetype = "dashed", 
             color = "deeppink",
             linewidth = 1.2) +
  geom_text(aes(x = breakpoint_date, y = 1, label = "SW", hjust = 1.5)) + 
  geom_vline(xintercept = mlt_cpt_date, 
             linetype = "dashed", 
             color = "darkblue",
             linewidth = 1.2) +
  geom_text(aes(x = mlt_cpt_date, y = 1, label = "MLT", hjust = 1.5))


```

Finally, a changepoint model for benthic vs pelagic dives.

```{r}
dive_summ <- read_rds(here::here("output/processed_dive_summaries.rds"))
dive_type <- read_csv(here::here("data/manual-dive-type.csv")) %>% 
  select(-c(AcousticsUsed, Notes)) %>% 
  mutate(ID = 1,
         Benthic_Pelagic_int = unclass(factor(Benthic_Pelagic, 
                                              levels = c("Benthic", 
                                                         "Pelagic"))),
         Rest_Forage_Travel_int = unclass(factor(Rest_Forage_Travel, 
                                                 levels = c("Rest", 
                                                            "Forage",
                                                            "Travel")))) 

dive_type_cpt <- dive_summ %>% 
  left_join(dive_type, by = "dive_id") %>% 
    drop_na(Benthic_Pelagic_int) %>% 
    mutate(date = as_date(start))

bp_cpt <- cpt.mean(dive_type_cpt$Benthic_Pelagic_int, method = "AMOC")
bp_cpt_date <- dive_type_cpt$date[bp_cpt@cpts[1]]

first(dive_type_cpt$date) - bp_cpt_date
last(dive_type_cpt$date) - bp_cpt_date
sum(dive_type_cpt$date < bp_cpt_date)
sum(dive_type_cpt$date > bp_cpt_date)

```

Show the MP plot with dive behavior below.

```{r}

daily_dives <- dive_summ %>% 
  left_join(dive_type, by = "dive_id") %>% 
  drop_na(Rest_Forage_Travel_int) %>% 
  mutate(date = lubridate::as_date(start)) %>% 
  group_by(date, Rest_Forage_Travel) %>% 
  summarize(count = n(), .groups = "drop_last") %>% 
  mutate(prop = count / sum(count)) %>% 
  ungroup()

p_divetype <- daily_dives %>% 
  ggplot(aes(date, prop, fill = Rest_Forage_Travel)) +
  geom_bar(stat = "identity", width = 0.8) + 
  scale_fill_manual(values = c("Rest" = "#1f78b4", 
                               "Forage" = "#a6cee3", 
                               "Travel" = "#b2df8a"
  )) +
  geom_vline(xintercept = as_date(breakpoint_date), 
             linetype = "dashed", 
             color = "deeppink",
             linewidth = 1.2) +
  geom_text(aes(x = as_date(breakpoint_date), y = 1, label = "SW"), hjust = 1.5) + 
  geom_vline(xintercept = as_date(mlt_cpt_date), 
             linetype = "dashed", 
             color = "darkblue",
             linewidth = 1.2) +
  geom_text(aes(x = as_date(mlt_cpt_date), y = 1, label = "MLT"), hjust = 1.5) +
  labs(x = NULL, 
       y = "Proportion of dives", 
       fill = "Behavior") + 
  theme_minimal()

cowplot::plot_grid(p_mp_cpt, p_divetype, nrow = 2, align ="h")

```

```{r}

phase_transitions <- as.POSIXct(c(
  "2000-01-01", # -Inf
  "2025-03-03",
  "2025-03-15",
  "2025-04-04",
  "2100-01-01"  # Inf
),
tz = "UTC")
step_length_km <- grab(fit, what = "predicted") %>% 
  mutate(dist_km = geosphere::distGeo(cbind(lon, lat)) / 1e3,
         phase = cut(date, 
                     phase_transitions, 
                     labels = c("Outbound", "Transition", "Foraging", "Inbound"))) 

step_length_km %>% 
  group_by(phase) %>% 
  summarize(mean_step_km = mean(dist_km, na.rm = TRUE), 
            sd_step_km = sd(dist_km, na.rm = TRUE),
            cv_step_km = sd_step_km / mean_step_km)

step_length_km %>% 
  ggplot(aes(date, dist_km)) +
  geom_line() +
  geom_point() +
  geom_rect(aes(xmin = start,
                xmax = end,
                ymin = -Inf,
                ymax = Inf),
            sw_valid,
            inherit.aes = FALSE,
            color = NA,
            fill = "pink",
            alpha = 0.5) +
  geom_vline(xintercept = breakpoint_date, 
             linetype = "dashed", 
             color = "deeppink",
             linewidth = 1.2) +
  geom_vline(xintercept = mlt_cpt_date, 
             linetype = "dashed", 
             color = "darkblue",
             linewidth = 1.2) +
  expand_limits(y = 0) 
  # coord_cartesian(xlim = c(breakpoint_date - 3600 * 3,
  #                          as.POSIXct("2025-03-15 06:00", tz = "UTC")))

```

```{r}
step_length_vect <- step_length_km %>% 
  filter(date >= as.POSIXct("2025-03-12", tz = "UTC")) %>% 
  vect(geom = c("lon", "lat"), crs = "EPSG:4326") %>% 
  project("+proj=tcea +lon_0=-124.7498444 +datum=WGS84 +units=m +no_defs") # Equal area proj

# Spatial range
c(x_range = ext(step_length_vect)[2] - ext(step_length_vect)[1],
  y_range = ext(step_length_vect)[4] - ext(step_length_vect)[3]) / 1e3

# centroid
geom(step_length_vect)[, c("x", "y")] %>% 
  colMeans() %>% 
  t() %>% 
  as_tibble() %>% 
  vect(geom = c("x", "y"), crs = crs(step_length_vect)) %>%
  project("EPSG:4326")

```
