---
title: "Dive behavior"
format: html
editor_options: 
  chunk_output_type: console
execute: 
  warning: false
---

Use `momentuHMM` to fit a Hidden Markov Model to the dive data.

First, read in and prep the data.

```{r}
#| label: setup
#| message: false

library(momentuHMM)
library(tidyverse)
source(here::here("R/dives.R"))

#read in dive type data (done manually)
dive_type <- read_csv(here::here("data/manual-dive-type.csv")) %>% 
  select(-c(AcousticsUsed, Notes)) %>% 
  mutate(ID = 1,
         Benthic_Pelagic_int = unclass(factor(Benthic_Pelagic, 
                                              levels = c("Benthic", 
                                                         "Pelagic"))),
         Rest_Forage_Travel_int = unclass(factor(Rest_Forage_Travel, 
                                                 levels = c("Rest", 
                                                            "Forage",
                                                            "Travel"))))

dive_prepped <- prepData(data = dive_type, 
                         coordNames = NULL)
```

Fit the HMM.

```{r}
#| label: fit-hmm

##### 3 state model #####

# Define initial parameters
nbStates <- 3  # number of hidden states

# order of initial states is Rest, Forage, Travel
one <- 0.99
zero <- function(k) (1 - one) / (k - 1)
cat_pars <- list(
  Rest_Forage_Travel_int = c(
    # P(R) for states R, F, T
    one, zero(3), zero(3),
    # P(F) for states R, F, T
    zero(3), one, zero(3)
    # P(T) implied
  )
)

# Fit HMM
hmm_fit3 <- fitHMM(data = dive_prepped,
                   nbStates = nbStates,
                   dist = list(Rest_Forage_Travel_int = "cat3"), 
                   Par0 = cat_pars,
                   stateNames = c("Rest", "Forage", "Travel"))
hmm_fit3
```

Now fitting the 3 state HMM with SW presence/absence in the dive as an environmental covariate.

```{r}

recording <- read_rds(here::here("output/recording_calib.rds"))
dive_summ <- read_rds(here::here("output/processed_dive_summaries.rds"))

###assign dive ids 
find_next <- function(x, y) {
  # Find the index of the next value in y following x
  approx(y, 
         seq_along(y),
         method = "constant",
         f = 1, 
         rule = 2,
         xout = x)$y
}

clicks_by_dive <- read_csv(here::here("data/sw/clicks_by_dive.csv")) %>% 
  #trying to connect recording with the correct dive id 
  mutate(end_calib = recording$end_calib[-nrow(recording)],
         dive_id = dive_summ$dive_id[find_next(end_calib, dive_summ$end)], 
         sw_yn = sw_yn == "y") %>% 
  select(-c(clickmanual_complete, buzz, notes)) %>% 
  left_join(select(dive_type, dive_id, Rest_Forage_Travel_int, ID),
            by = "dive_id")

dive_sw_prepped <- prepData(clicks_by_dive, 
                            coordNames = NULL)

# Fit HMM w sw as a covariate
hmm_fit3_sw <- fitHMM(data = dive_sw_prepped,
                      nbStates = nbStates,
                      formula = ~ sw_yn,
                      dist = list(Rest_Forage_Travel_int = "cat3"), 
                      Par0 = cat_pars,
                      fixPar = list(beta = matrix(
                        c(NA, NA, NA, NA, NA, NA,
                          NA,  0, NA,  0,  0,  0),
                        nrow = 2, byrow = TRUE
                      )),
                      stateNames = c("Rest", "Forage", "Travel"))

hmm_fit3_sw

ci <- CIbeta(hmm_fit3_sw, alpha = 0.05)
ci$beta  # coefficients, SEs, and 95% CIs for each transition
#some NAs for transitions where probabilities were constrained near 0. 

beta_table <- bind_cols(
  as.data.frame.table(ci$beta$est,   
                      responseName = "est"),
  as.data.frame.table(ci$beta$se,    
                      responseName = "se") %>%  
    select(se),
  as.data.frame.table(ci$beta$lower, 
                      responseName = "lower") %>% 
    select(lower),
  as.data.frame.table(ci$beta$upper, 
                      responseName = "upper") %>%  
    select(upper)
) %>% 
  rename(parameter = Var1, transition = Var2) 

library(kableExtra)
beta_table %>% 
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>% #round to 3 places
kable(col.names = c("Parameter", "Transition", "Estimate", "SE", "Lower 95%", "Upper 95%"),
        caption = "Beta coefficients") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) 
```

If the CI excludes zero, the covariate effect on that transition is significant.

```{r}

# dive_sw_prepped %>%
#   mutate(state = Rest_Forage_Travel_int,
#          next_state = lead(Rest_Forage_Travel_int)) %>% 
#   count(state, next_state, sw_yn) %>% 
#   drop_na(next_state) %>% 
#   group_by(state, sw_yn) %>% 
#   mutate(freq = n / sum(n)) %>% 
#   ungroup() %>% 
#   arrange(state, sw_yn, next_state) %>% 
#   filter(next_state == 2) %>% 
#   view()

p <- ggplot(dive_sw_prepped, aes(start_time, Rest_Forage_Travel_int)) +
  geom_line() +
  geom_point(aes(color = sw_yn, group = dive_id))
plotly::ggplotly(p, dynamicTicks = TRUE)

```

Are there differences in the dive characteristics when sw are present?

```{r}

dive_sw_prepped %>% 
  group_by(Rest_Forage_Travel_int, sw_yn) %>% 
  summarize(dur_mean = mean(duration_min),
            dur_sd = sd(duration_min),
            .groups = "drop") %>% 
  mutate(sw_yn = ifelse(sw_yn, "sw_present", "sw_absent")) %>% 
  pivot_wider(names_from = sw_yn, values_from = c(dur_mean, dur_sd)) 

library(nlme)
# Limit analysis to foraging dives 10+ minutes long
foraging_dives <- dive_sw_prepped %>% 
  filter(duration_min >= 10, 
         Rest_Forage_Travel_int == 2) %>% 
  mutate(hours_elapsed = as.numeric(start_time - start_time[1], unit = "hours"))
dive_dur <- gls(duration_min ~ sw_yn, 
                foraging_dives,
                correlation = corCAR1(form = ~ hours_elapsed))
summary(dive_dur)

# binned ACF for each model
irreg_acf <- function(mod) {
  r <- residuals(mod, type = "normalized")
  t <- foraging_dives$hours_elapsed
  
  pairs <- expand_grid(i = seq_along(r), j = seq_along(r)) %>% 
    filter(j > i) %>% 
    mutate(lag = abs(t[i] - t[j]),
           prod = r[i] * r[j]) %>% 
    filter(lag <= 24)
  
  bin_width = 0.5
  pairs %>% 
    mutate(lag_bin = floor(lag / bin_width) * bin_width + bin_width / 2) %>% 
    group_by(lag_bin) %>% 
    summarize(acf = mean(prod), n = n()) %>% 
    slice(-n()) %>% 
    ggplot(aes(lag_bin, acf)) +
    geom_hline(yintercept = 0 ) +
    geom_hline(yintercept = c(-1.96, 1.96) / sqrt(length(t)),
               linetype = "dashed", color = "blue") +
    geom_col()
}
irreg_acf(dive_dur)

```

Dive duration is on average 1-2 mins longer when sperm whales are present.
