## Broad scale

Exploring the `aniMotum` package as a way to fit a continuous HMM to the seal GPS data.

I followed the installation instructions [here](https://ianjonsen.github.io/aniMotum/index.html).

```{r}
#| label: setup
#| message: false

library(aniMotum)
library(tidyverse)
library(betareg)

theme_set(theme_bw())
```

Loading in our GPS data and formatting it to work with `aniMotum`:

```{r}
#| label: data

track <- read_csv(here::here("data/seal_locations.csv"),
                  show_col_types = FALSE) %>% 
  mutate(lc = c("G"), 
         id = c("H391")) %>% 
  format_data(date = "Date_es", coord = c("Longitude", "Latitude")) %>% 
  as.data.frame()

```

The general process for fitting movement persistence models with `aniMotum` is to use the `fit_ssm` function with the argument `model = "mp"` (i.e., movement persistence).

```{r}

fit <- fit_ssm(track, 
               vmax = 3, #max travel rate 
               model = "mp", #fits continuous HMM for movement persistence
               time.step = 24, #daily steps
               control = ssm_control(verbose = 0)) #turns off reports

summary(fit)
```

*Note: I chose `vmax = 3` because the example (also on elephant seals) chose that. It would be good to double check that this makes sense.*

```{r}

plot(fit, what = "predicted", type = 3, normalise = TRUE)
```

Map:

```{r}
aniMotum::map(fit, what = "p", normalise = TRUE, silent = TRUE)
```

AniMotum is giving us persistence, but I can't include the covariate of odontocete detections. However, this does confirm that persistence as a metric describes the broad-scale behavioral states. Next up, plot the persistence metric ( $\gamma$ ) as a function of "odontocete intensity" (number of detections). Max says use beta regression as an exploratory analysis.

Load in the exploratory acoustic data.

```{r}

acou <- read_csv(here::here("data/acoustic-explore.csv"))

page_start <- as.POSIXct("2025-02-21 06:00", tz = "US/Pacific")
window_hr <- 24
find_window <- function(time, start, win_hr) {
  n_window <- as.numeric(time - start, unit = "hours") %/% win_hr
  start + n_window * window_hr * 3600 + window_hr / 2 * 3600
}
acou_agg <- acou %>% 
  mutate(time = page_start + (page - 1) * 3600,
         date = find_window(time, page_start, window_hr)) %>% 
  group_by(date) %>% 
  summarize(n_odon = sum(odontocetes)) %>% 
  mutate(date = lubridate::with_tz(date, "Etc/UTC"))

```

Plot $\gamma$ as a function of the number of odontocete detections.

```{r}

gamma <- fit$ssm$H391$predicted %>% select(date, g) %>% mutate(date = date - 3600)

acou_gamma <- left_join(acou_agg, gamma, by = "date") %>% replace_na(list(n_odon= 0))

ggplot(acou_gamma, aes(n_odon, g)) + 
  geom_path() +
  geom_point(aes(color = date)) + 
  scale_color_distiller(palette = "RdYlBu")

betareg()
```

Lower left corner: those points had odontocetes on preceding and subsequent days. This has removed the fact that this was a continuous process, so the relationship is lost. We need to model this as a temporal process, and also account for the fact that different odontocetes have different effects.
