---
title: Simulate Continuous-time Hidden Markov Model
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(CircStats)
theme_set(theme_bw())

```

## What is a CTHMM?

It's a HMM but the observations come at inconsistent intervals. Therefore estimate state transition *rates* instead of *probabilities*.

## Seal behavior

At the coarse scale, two states: transient and resident. Transient state is characterized by longer step lengths and more concentrated turning angles; resident the opposite. Assume step lengths and turning angles are distributed as gamma and von Mises distributions, respectively.

```{r}
#| label: fig-ideal-emissions

step_param <- list(
  transient = c(mean = 25, dispersion = 0.2),
  resident = c(mean = 3, dispersion = 0.15)
)

expand_grid(
  state = c("transient", "resident"),
  step_length = seq(0.01, 50, length.out = 1000)
) %>% 
  mutate(step_mean = map_dbl(state, \(s) step_param[[s]]["mean"]),
         step_disp = map_dbl(state, \(s) step_param[[s]]["dispersion"]),
         # See below about mean/dispersion -> shape/scale
         step_var = step_disp * step_mean^2,
         step_dens = dgamma(step_length, 
                            shape = step_mean^2 / step_var, 
                            rate = step_mean / step_var)) %>% 
  ggplot(aes(step_length, step_dens, color = state)) +
  geom_line()

# By default, gamma is parameterized using SHAPE and RATE or SHAPE and SCALE.
# Neither of these parameters are intuitive, so we pick a MEAN and DISPERSION
# instead, then apply a formula to convert them to SHAPE and RATE.

```
